---
title: "lingtypology"
author: "G. Moroz"
output: html_document
---

lingtypology provides R with the glottolog database http://glottolog.org/ (v. 2.7) and some more abilities for purposes of linguistic typology. The glottolog database containes the catalogue of languages of the world. This package helps researchers to make a linguistic maps, using philosophy of The Cross-Linguistic Linked Data project (http://clld.org/), which allows for while at the same time facilitating uniform access to the data across publications.

## Installation

Get the released version from github:

```{r}
install.packages("devtools")
devtools::install_github("agricolamz/lingtypology")
```

погнали... потом переведу
## Базовые функции
В пакет встроены фрагменты базы данных Glottolog: языкоиды (linguoids), ISO коды, координаты, аффилиация и страны распространения языков. Для того, чтобы даставать эту информацию в пакете lingtypology есть несколько функций, имеющих структуру дай мне X на основе Y.
```{r}
library(lingtypology)
lang.iso("Adyghe")
```
