---
title: "`lingtypology`: Glottolog functions"
author: "George Moroz"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{`lingtypology`: Glottolog functions}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
library(lingtypology)
```

This package is based on the [Glottolog database (v. 4.4)](https://glottolog.org/), so `lingtypology` has several functions for accessing data from that database.

### 1. Command name's syntax
Most of the functions in `lingtypology` have the same syntax: **what you need.what you have**. Most of them are based on _language name_.

* **aff.lang()** — get affiliation by language
* **area.lang()** — get macro area by language
* **country.lang()** — get country by language
* **iso.lang()** — get ISO 639-3 code by language
* **gltc.lang()** — get glottocode (identifier for a language in the Glottolog database) code by language
* **lat.lang()** — get latitude by language
* **long.lang()** — get longitude by language
* **level.lang()** — get level by language
* **subc.lang()** — get subclassification in the Newick tree format by language

Some of them help to define a vector of languages.

* **lang.aff()** — get language by affiliation
* **lang.iso()** — get language by ISO 639-3 code
* **lang.gltc()** — get language by glottocode

Additionally there are some functions to convert glottocodes to ISO 639-3 codes and vice versa:

* **gltc.iso()** — get glottocode by ISO 639-3 code
* **iso.gltc()** — get ISO 639-3 code by glottocode

The most important functionality of `lingtypology` is the ability to create interactive maps based on features and sets of languages (see the third section):

* **map.feature()**

[Glottolog database (v. 4.1)](https://glottolog.org/) provides `lingtypology` with language names, ISO codes, glottocodes, affiliation, macro area, coordinates, and much information. This set of functions doesn't have a goal to cover all possible combinations of functions. Check out additional information that is preserved in the version of the Glottolog database used in `lingtypology`:

```{r}
names(glottolog)
```

Using R functions for data manipulation you can create your own database for your purpose.

### 2. Using base functions
All functions introduced in the previous section are regular functions, so they can take the following objects as input:

* a regular string
```{r}
iso.lang("Adyghe")
lang.iso("ady")
lang.aff("West Caucasian")
```

I would like to point out that you can create strings in R using single or double quotes. Since inserting single quotes in a string created with single quotes causes an error in R, I use double quotes in my tutorial. You can use single quotes, but be careful and remember that `'Ma'ya'` is an incorrect string in R.

* a vector of strings
```{r}
area.lang(c("Adyghe", "Aduge"))
lang <- c("Adyghe", "Russian")
aff.lang(lang)
```

*  other functions. For example, let's try to get a vector of ISO codes for the Circassian languages
```{r}
iso.lang(lang.aff("Circassian"))
```

If you are new to R, it is important to mention that you can create a table with languages, features and other parametres with any spreadsheet software you used to work. Then you can import the created file to R using standard tools.

### 3. Spell Checker: look carefully at warnings!

All functions which take a vector of languages are enriched with a kind of a spell checker. If a language from a query is absent in the database, functions return a warning message containing a set of candidates with the minimal Levenshtein distance to the language from the query.

```{r}
aff.lang("Adyge")
```

### 4. `subc.lang()` function

The `subc.lang()` function returns language subclassification in the Newick tree format.

```{r}
subc.lang("Lechitic")
```

This format is hard to interpret by itself, but there are some tools in R that make it possible to visualise those subclassifications:

```{r}
library(ape)
plot(read.tree(text = subc.lang("Lechitic")))
```

It is possible to specify colors of tips in case you want to emphasize some nodes:

```{r}
plot(read.tree(text = subc.lang("Lechitic")),
     tip.color = c("red", "black", "black", "black"))
```

As you can see nodes are counted from bottom to top.

For more sophisticated tree visualization you can look into [`ggtree` package](https://bioconductor.org/packages/release/bioc/html/ggtree.html). There are several linguistic packages that provide some functionality for creating glottolog trees:

* [`glottoTrees`](https://github.com/erichround/glottoTrees) package by Erich Round
* [`lingtypr`](https://gitlab.com/laurabecker/lingtypr) package by Laura Becker
