---
title: "`lingtypology`: easy mapping for Lingustic Typology"
author: "George Moroz"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{`lingtypology`: easy mapping for Lingustic Typology}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## What is lingtypology?
The `lingtypology` package connects R with the [Glottolog database (v. 2.7)](http://glottolog.org/) and provides an additional functionality for linguistic typology. The Glottolog database contains a catalogue of the world's languages. This package helps researchers to make linguistic maps, using the philosophy of [the Cross-Linguistic Linked Data project](http://clld.org/), which is creating uniform access to linguistic data across publications. This package is based on the [leaflet package](https://rstudio.github.io/leaflet/), so `lingtypology` is a package for interactive linguistic cartography. I would like to thank Natalya Tyshkevich and Samira Verhees for reading and correcting this vignette.

## 1. Installation
Since `lingtypology` is an R package, you should install [R](https://www.r-project.org/) on your PC if you haven't already done so. To install the `lingtypology` package, run the following command at your R IDE, so you get the stable version from CRAN:
```{r, eval=FALSE}
install.packages("lingtypology")
```

You can also get the development version from GitHub:
```{r, eval= F}
install.packages("devtools")
devtools::install_github("agricolamz/lingtypology")
```

Load package:
```{r}
library(lingtypology)
```

## 2. Glottolog functions
This package is based on the [Glottolog database (v. 2.7)](http://glottolog.org/), so `lingtypology` has several functions for accessing data from that database. In the Glottolog database, the term [_languoid_](http://glottolog.org/glottolog/glottologinformation) is used to catalogue languages, dialects and language families alike.

### 2.1 Command name's syntax
Most of the functions in `lingtypology` have the same syntax: **what you need.what you have**. Most of them are based on _languoid name_.

* **aff.lang()** — get affiliation by languoid
* **area.lang()** — get macro area by languoid
* **country.lang()** — get country by languoid
* **iso.lang()** — get ISO 639-3 code by languoid
* **lat.lang()** — get latitude by languoid
* **long.lang()** — get longitude by languoid

Some of them help to define a vector of languoids.

* **lang.aff()** — get languoid by affiliation
* **lang.iso()** — get languoid by ISO 639-3 code

The most important functionality of `lingtypology` is the ability to create interactive maps based on features and sets of languoids (see the third section):

* **map.feature()**

[Glottolog database (v. 2.7)](http://glottolog.org/) provides `lingtypology` with  languoid names, ISO codes, genealogical affiliation, macro area, countries and coordinates.

### 2.2 Using base functions
All functions introduced in the previous section are regular functions, so they can take the following objects as input:

* a regular string
```{r}
iso.lang("Adyghe")
lang.iso("ady")
country.lang("Adyghe")
lang.aff("Abkhaz-Adyge")
```
I would like to point out that strings in R can be created using
single or double quotes. Since inserting single quotes in a string created with single quotes causes an error in R, I use double quotes in my tutorial. You can use single quotes, but be careful and remember that `'Ma'ya'` is an incorrect string in R.

* a vector of strings
```{r}
area.lang(c("Adyghe", "Aduge"))
lang <- c("Adyghe", "Russian")
aff.lang(lang)
```
*  other functions. For example, let's try to get vector of ISO codes for the East Slavic languages
```{r}
iso.lang(lang.aff("East Slavic"))
```
The behavior of most functions is rather predictable, but the function `country.lang`  has an additional feature. By default this function takes a vector of languages and returns a vector of countries.  But if you set the argument `intersection = TRUE`,  then the function returns a vector of countries where all languoids from the query are spoken.
```{r}
country.lang(c("Udi", "Laz"))
country.lang(c("Udi", "Laz"), intersection = TRUE)
```
### 2.3 Spell Checker: look carefully at warnings!
There are some functions that take country names as input. Unfortunately, some countries have alternative names. In order to save users the trouble of having to figure out the exact name stored in the database (for example  _Ivory Coast_ or _Cote d'Ivoire_), all official country names and standard abbreviations are stored in the database:

```{r}
lang.country("Cape Verde")
lang.country("Cabo Verde")
head(lang.country("UK"))
```
All functions which take a vector of languoids are enriched with a kind of a spell checker. If a languoid from a query is absent in the database, functions return a warning message containing a set of candidates with the minimal Levenshtein distance to the languoid from the query.
```{r}
aff.lang("Adyge")
```

### 2.4 Changes in the glottolog database
Unfortunately, the [Glottolog database (v. 2.7)](http://glottolog.org/) is not perfect for all my tasks, so I changed it a little bit:

* Added some Dargwa languages: Chirag, Cudaxar, Itsari, Kajtak, Kubachi, Tanti, Mehweb. Dargwa is still present in the database as an individual languoid, so one can use Dargwa or any Dargwa language listed above. (07.09.2016)
* Tabassaran is renamed Tabasaran (07.09.2016)
* Added coordinates to Silesian and Slavomolisano (11.09.2016)
* North Azerbaijani is renamed Azerbaijani. Added some Dargwa languages: Amuzgi-Shiri, Gapshima, North-Dargwa, Verhnevurkunskij. Added some Avar dialects: Avar (Andalal), Avar (Antsukh), Avar (Batlukh), Avar (Gid), Avar (Karah), Avar (Karata), Avar (Zakatal)  (11.12.2016)
* Sauk (Trans New Guinea) is renamed Ma Manda. (27.12.2016)
* Ache is renamed Ache (Sino Tibetian) and Ache (Tupian); Bari is renamed Bari (Chibchan) and Bari (Nilotic); Bugan is renamed Bugan and Bugan (Spurious); Calo is renamed Calo and Calo (Spurious); Chimakum is renamed Chimakum and Chimakum (Spurious); Garawa is renamed Garawa and Garawa (Spurious); Hamba is renamed Hamba (Atlantic-Congo) and Hamba (Unclassifiable); Hoti is renamed Hoti (Jodi-Saliban) and Hoti (Austronesian); Karipuna is renamed Karipuna (Pano-Tacanan) and Karipuna (Tupian); Kula is renamed Kula (Austronesian) and Kula (Timor-Alor-Pantar); Laru is renamed Laru (Atlantic-Congo) and Laru (Heiban); Maring is renamed Maring (Nuclear Trans New Guinea) and Maring (Sino-Tibetan); Mayo is renamed Mayo (Uto-Aztecan) and Mayo (Pano-Tacanan); Nabi is renamed Nabi (Austronesian) and Nabi (Nuclear Torricelli); Natugu is renamed Natugu and Natugu (Spurious); Naxi is renamed Naxi and Naxi (Spurious); Ngen is renamed Ngen (Mande) and Ngen (Austronesian); Purum is renamed Purum and Purum (Spurious); Saliba is renamed Saliba (Austronesian) and Saliba (Jodi-Saliban); Southern Betsimisaraka Malagasy is renamed Southern Betsimisaraka Malagasy and Southern Betsimisaraka Malagasy (Spurious); Surigaonon is renamed Surigaonon and Surigaonon (Spurious); Tai Pao is renamed Tai Pao and Tai Pao (Spurious); Tunen is renamed Tunen and Tunen (Spurious); Wara is renamed Wara (Atlantic-Congo) and Wara (Morehead-Wasur); Yendang is renamed Yendang and Yendang (Spurious) (22.01.2017)
* 360+ longitude conversion made for Austral, Central Siberian Yupik, East Futuna, Hawaiian, Lauan, Mangarevam Naukan Yupik, Niuafo'ou, Niuatoputapu, Niuean, North Marquesan, Old Rapa, Penrhyn, Pitcairn-Norfolk, Pukapuka, Rakahanga-Manihiki, Rapanui, Rarotongan, Samoan, Sirenik Yupik, South Marquesan, Tahitian, Tokelau, Tonga (Tonga Islands), Tuamotuan, Wallisian (22.01.2017)
* Abkhazian is changed to Abkhaz. Abkhazian, Abaza and Georgian coordinates are changed. (22.01.2017)
* Northern Sahaptin names corrected. Central Guerrero Nahuatl is renamed Guerrero Nahuatl (thanks to Timo Roettger for mentioning that problems). (02.02.2017)
* Some sign language changed. Thanks to [Calle Börstell](https://borstell.github.io/#name) for data. (02.02.2017)
* Bininj Gun-Wok changed to Gunwinggu (thanks to Timo Roettger) (06.02.2017)
* Country name Viet Nam changed to Vietnam. Madagascar and Togo added to countries set. (11.02.2017)

After [Robert Forkel's issue](https://github.com/agricolamz/lingtypology/issues/1) I decided to add an argument `glottolog.source`, so that everybody has access to "original" and "modified" (by default) glottolog versions:
```{r}
is.glottolog(c("Tabasaran", "Tabassaran"), glottolog.source = "original")
```
```{r}
is.glottolog(c("Tabasaran", "Tabassaran"), glottolog.source = "modified")
```

It is common practice in R to reduce both function arguments and its values, so this can also be done with the following lingtypology functions.
```{r}
is.glottolog(c("Tabasaran", "Tabassaran"), g = "o")
```
```{r}
is.glottolog(c("Tabasaran", "Tabassaran"), g = "m")
```


## 3. Map features with `map.feature`
### 3.1 Base map
The most important part of the `lingtypology` package is the function `map.feature`. This function allows a user to produce maps similar to known projects within [the Cross-Linguistic Linked Data philosophy](http://clld.org/), such as [WALS](http://wals.info/) and [Glottolog](http://glottolog.org/):
```{r, fig.width=6.2}
map.feature(c("Adyghe", "Kabardian", "Polish", "Russian", "Bulgarian"))
```

As shown in the picture above, this function generates an interactive Leaflet map with a control box that allows users to toggle the visibility of any group of points on the map. All specific points on the map have a pop-up box that appears when markers are clicked (see section 3.3 for more information about editing pop-up boxes). By default, they contain languoid names linked to the glottolog site.

If you are new to R, please find some information about how to import data to R. It is simple to make a .csv, .ods or .xls files containing lists of languages and features and read it from R (.csv is the easiest way).

If for some reasons you are not using RStudio or you want to automatically create and save a lot of maps, you can save a map to a variable and use the `htmlwidgets` package for saving created maps to an .html file. I would like to thank Timo Roettger for mentioning this problem.

```{r, eval = FALSE}
m <- map.feature(c("Adyghe", "Korean"))
# install.packages("htmlwidgets")
library(htmlwidgets)
saveWidget(m, file="/home/agricolamz/_DATA/OneDrive1/_Work/github/lingtypology/m.html")
```

There is an export button in RStudio, but for some reason it is not so easy to save a map as a .png or .jpg file using code. [Here](http://stackoverflow.com/a/34672309/6056442) is a possible solution.

### 3.2 Set features
The goal of this package is to allow typologists (or any other linguists) to map language features. A list of languoids and correspondent features can be stored in a `data.frame` as follows:
```{r}
df <- data.frame(language = c("Adyghe", "Kabardian", "Polish", "Russian", "Bulgarian"),
                 features = c("polysynthetic", "polysynthetic", "fusional", "fusional", "fusional"))
df
```
Now we can draw a map:
```{r, fig.width=6.2}
map.feature(languages = df$language, features = df$features)
```

If you have a lot of features and they appear in the legend in a senseless order (by default it is ordered alphabetically), you can reorder them using factors (a vector with ordered levels, for more information see `?factor`). For example, I want the feature polysynthetic to be listed first, followed by fusional:

```{r, fig.width=6.2}
df$features <- factor(df$features, levels = c("polysynthetic", "fusional"))
map.feature(languages = df$language, features = df$features)
```

Since the correspondence between a color palette and a mapped features is chosen randomly by default, it is better to use the function `set.seed`, to get a reproducible map (or choose colors yourself, see section 3.5):
```{r, fig.width=6.2}
set.seed(42)
map.feature(languages = df$language, features = df$features)
```


Like in most R functions, it is not necessary to name all arguments, so the same result can be obtained by:
```{r, fig.width=6.2}
set.seed(42)
map.feature(df$language, df$features)
```
As shown in the picture above, all points are grouped by feature, colored and counted. As before, a pop-up box appears when markers are clicked. A control feature allows users to toggle the visibility of points grouped by feature.

### 3.3 Set pop-up boxes
Sometimes it is a good idea to add some additional information (e.g. language affiliation,  references or even examples) to pop-up boxes that appear when points are clicked. In order to do so, first of all we need to create an extra vector of strings in our dataframe:
```{r, fig.width=6.2}
df$popup <- aff.lang(df$language)
```

The function `aff.lang()`  creates a vector of genealogical affiliations that can be easily mapped:
```{r, fig.width=6.2}
set.seed(42)
map.feature(languages = df$language, features = df$features, popup = df$popup)
```

Like before, it is not necessary to name all arguments, so the same result can be obtained by this:
```{r, fig.width=6.2}
set.seed(42)
map.feature(df$language, df$features, df$popup)
```

Pop-up strings can contain HTML tags, so it is easy to insert a link, a couple of lines, a table or even a video and sound. Here is how pop-up boxes can demonstrate language examples:
```{r, fig.width=6.2}
# change a df$popup vector
df$popup <- c ("sɐ s-ɐ-k'ʷɐ<br> 1sg 1sg.abs-dyn-go<br>'I go'",
               "sɐ s-o-k'ʷɐ<br> 1sg 1sg.abs-dyn-go<br>'I go'",
               "id-ę<br> go-1sg.npst<br> 'I go'",
               "ya id-u<br> 1sg go-1sg.npst <br> 'I go'",
               "id-a<br> go-1sg.prs<br> 'I go'")
# create a map
set.seed(42)
map.feature(df$language, df$features, df$popup)
```

How to say _moon_ in Sign Languages? Here is an example:
```{r, fig.width=6.2}
# Lets create a dataframe with links to video
sign_df <- data.frame(languages = c("American Sign Language", "Russian Sign Language", "French Sign Language"),
                 popup = c("https://media.spreadthesign.com/video/mp4/13/48600.mp4", "https://media.spreadthesign.com/video/mp4/12/17639.mp4", "https://media.spreadthesign.com/video/mp4/10/17638.mp4"))

# Change popup to an HTML code
sign_df$popup <- paste("<video width='200' height='150' controls> <source src='",
                  as.character(sign_df$popup),
                  "' type='video/mp4'></video>", sep = "")

# create a map
map.feature(languages = sign_df$languages, popup = sign_df$popup)
```



### 3.4 Set labels
An alternative way to add some short text to a map is to use the `label` option.
```{r, fig.width=6.2}
set.seed(42)
map.feature(df$language, df$features,
            label = df$language)
```

There are some additional arguments for customization: `label.fsize` for setting font size, `label.position` for controlling the label position, and `label.hide` to control the appearance of the label: if `TRUE`, the labels are displayed on mouse over (as on the previous map), if `FALSE`, the labels are always displayed (as on the next map).

```{r, fig.width=6.2}
set.seed(42)
map.feature(df$language, df$features,
            label = df$language,
            label.fsize = 20,
            label.position = "left",
            label.hide = TRUE)
```



### 3.5 Set coordinates
Users can set their own coordinates using the arguments `latitude` and `longitude`. I will illustrate this with the dataset `circassian` built into the `lingtypology` package. This dataset comes from fieldwork collected during several expeditions in the period 2011-2016 and contains a list of Circassian villages:
```{r, fig.width=6.2}
set.seed(42)
map.feature(languages = circassian$language,
            features = circassian$languoid,
            popup = circassian$village,
            latitude = circassian$latitude,
            longitude = circassian$longitude)
```

### 3.6 Set colors
By default the color palette is created by the `rainbow()` function, but users can set their own colors using the argument `color`:
```{r, fig.width=6.2}
df <- data.frame(language = c("Adyghe", "Kabardian", "Polish", "Russian", "Bulgarian"),
                 features = c("polysynthetic", "polysynthetic", "fusional", "fusional", "fusional"))
map.feature(languages = df$language,
            features = df$features,
            color = c("yellowgreen", "navy"))
```

### 3.7 Set control box
The package can generate a control box that allows users to toggle the visibility of points and features. To enable it, there is an argument  `control` in the `map.feature` function:

```{r, fig.width=6.2}
set.seed(42)
map.feature(languages = df$language,
            features = df$features,
            control = TRUE)
```

### 3.8 Set an additional set of features using strokes
The `map.feature` function has an additional argument `stroke.features`. Using this argument it becomes possible to show two independent sets of features on one map. By default strokes are colored in grey (so for two levels it will be black and white, for three --- black, grey, white end so on), but users can set their own colors using the argument `stroke.color`:
```{r, fig.width=6.2}
set.seed(42)
map.feature(circassian$language,
            features = circassian$languoid,
            stroke.features = circassian$language,
            latitude = circassian$latitude,
            longitude = circassian$longitude)
```

It is important to note that `stroke.features` can work with `NA` values. The function won't plot anything if there is an `NA` value. Let's set a language value to `NA` in all Baksan villages from the `circassian` dataset
```{r, message= F, fig.width=6.2}
# create newfeature variable
newfeature <- circassian[,c(5,6)]
# set language feature of the Baksan villages to NA and reduce newfeature from dataframe to vector
newfeature <-  replace(newfeature$language, newfeature$languoid == "Baksan", NA)
# create a map
set.seed(42)
map.feature(circassian$language,
            features = circassian$languoid,
            latitude = circassian$latitude,
            longitude = circassian$longitude,
            stroke.features = newfeature)
```

### 3.9 Set radii and an opacity feature
All markers have their own radius and opacity, so it can be set by users. Just use the arguments `radius`, `stroke.radius`, `opacity` and `stroke.opacity`:
```{r, fig.width=6.2}
set.seed(42)
map.feature(circassian$language,
            features = circassian$languoid,
            stroke.features = circassian$language,
            latitude = circassian$latitude,
            longitude = circassian$longitude,
            radius = 7, stroke.radius = 13)

set.seed(42)
map.feature(circassian$language,
            features = circassian$languoid,
            stroke.features = circassian$language,
            latitude = circassian$latitude,
            longitude = circassian$longitude,
            opacity = 0.7, stroke.opacity = 0.6)
```

### 3.10 Customizing legends
By default the legend appears in the bottom left corner. If there are stroke features, two legends are generated. There are additional arguments that control the appearence and the title of the legends.
```{r, fig.width=6.2}
set.seed(42)
map.feature(circassian$language,
            features = circassian$languoid,
            stroke.features = circassian$language,
            latitude = circassian$latitude,
            longitude = circassian$longitude,
            legend = FALSE, stroke.legend = TRUE)

set.seed(42)
map.feature(circassian$language,
            features = circassian$languoid,
            stroke.features = circassian$language,
            latitude = circassian$latitude,
            longitude = circassian$longitude,
            title = "Circassian dialects", stroke.title = "Languages")
```
The arguments `legend.position` and `stroke.legend.position` allow users to change a legend's position using "topright", "bottomright", "bottomleft" or "topleft" strings.

### 3.11 Set scale bar
A scale bar is automatically added to a map, but users can control its appearance (set `scale.bar` argument to `TRUE` or `FALSE`) and its position (use `scale.bar.position` argument values "topright", "bottomright", "bottomleft" or "topleft").
```{r, fig.width=6.2}
set.seed(42)
map.feature(c("Adyghe", "Polish", "Kabardian", "Russian"),
            scale.bar = TRUE,
            scale.bar.position = "topright")
```

### 3.12 Set layouts
It is possible to use different tiles on the same map  using the `tile` argument. For more tiles see [here](https://leaflet-extras.github.io/leaflet-providers/preview/index.html).
```{r, fig.width=6.2}
df <- data.frame(lang = c("Adyghe", "Kabardian", "Polish", "Russian", "Bulgarian"),
   feature = c("polysynthetic", "polysynthetic", "fusion", "fusion", "fusion"),
   popup = c("Adyghe", "Adyghe", "Slavic", "Slavic", "Slavic"))
set.seed(42)
map.feature(df$lang, df$feature, df$popup,
            tile = "Thunderforest.OpenCycleMap")
```

It is possible to use different map tiles on the same map. Just add a vector with tiles.
```{r, fig.width=6.2}
df <- data.frame(lang = c("Adyghe", "Kabardian", "Polish", "Russian", "Bulgarian"),
   feature = c("polysynthetic", "polysynthetic", "fusion", "fusion", "fusion"),
   popup = c("Adyghe", "Adyghe", "Slavic", "Slavic", "Slavic"))
set.seed(42)
map.feature(df$lang, df$feature, df$popup,
            tile = c("OpenStreetMap.BlackAndWhite", "Thunderforest.OpenCycleMap"))
```

It is possible to name tiles using the `tile.name` argument.
```{r, fig.width=6.2}
df <- data.frame(lang = c("Adyghe", "Kabardian", "Polish", "Russian", "Bulgarian"),
   feature = c("polysynthetic", "polysynthetic", "fusion", "fusion", "fusion"),
   popup = c("Adyghe", "Adyghe", "Slavic", "Slavic", "Slavic"))
set.seed(42)
map.feature(df$lang, df$feature, df$popup,
            tile = c("OpenStreetMap.BlackAndWhite", "Thunderforest.OpenCycleMap"),
            tile.name = c("b & w", "colored"))
```

It is possible to combine the tiles' control box with the features' control box.
```{r, fig.width=6.2}
df <- data.frame(lang = c("Adyghe", "Kabardian", "Polish", "Russian", "Bulgarian"),
   feature = c("polysynthetic", "polysynthetic", "fusion", "fusion", "fusion"),
   popup = c("Adyghe", "Adyghe", "Slavic", "Slavic", "Slavic"))
set.seed(42)
map.feature(df$lang, df$feature, df$popup,
            tile = c("OpenStreetMap.BlackAndWhite", "Thunderforest.OpenCycleMap"),
            control = TRUE)
```

### 3.13 Add a minimap to a map
It is possible to add a minimap to a map.
```{r, fig.width=6.2}
set.seed(42)
map.feature(c("Adyghe", "Polish", "Kabardian", "Russian"),
            minimap = TRUE)
```

Users can control its appearance (by setting the `minimap` argument to TRUE or FALSE), its position (by using the values "topright", "bottomright", "bottomleft" or "topleft" of the `minimap.position` argument) and its height and width (with the arguments `minimap.height` and `minimap.width`).
```{r, fig.width=6.2}
set.seed(42)
map.feature(c("Adyghe", "Polish", "Kabardian", "Russian"),
            minimap = TRUE,
            minimap.position = "topright",
            minimap.height = 100,
            minimap.width = 100)
```


### 3.14 Add a picture to a map
The argument images.url allows users to add their own pictures to a map, using an url. In this part I will use two histograms on the most numerous nationalities in Moscow and St. Petersburg, based on data from the last Russian Census:

* [Moscow](http://web-corpora.net/~agricolamz/maps/lingtypology/Moscow.png)
* [St. Petersburg](http://web-corpora.net/~agricolamz/maps/lingtypology/St.%20Petersburg.png)

Let's create a dataframe.
```{r, fig.width=6.2}
df <- data.frame(lang = c("Russian", "Russian"),
                 lat  = c(55.75, 59.95),
                 long = c(37.616667, 30.3),
# I use here URL shortener by Google
                 urls = c("https://goo.gl/5OUv1E",
                          "https://goo.gl/UWmvDw"))
map.feature(languages = df$lang,
            latitude = df$lat,
            longitude = df$long,
            image.url = df$urls)
```

Users can change the size of the pictures.
```{r, fig.width=6.2}
df <- data.frame(lang = c("Russian", "Russian"),
                 lat  = c(55.75, 59.95),
                 long = c(37.616667, 30.3),
# I use here URL shorter by Google
                 urls = c("https://goo.gl/5OUv1E",
                          "https://goo.gl/UWmvDw"))
map.feature(languages = df$lang,
            latitude = df$lat,
            longitude = df$long,
            image.url = df$urls,
            image.width = 200,
            image.height = 200)
```

It can be moved from the actual point:
```{r, fig.width=6.2}
df <- data.frame(lang = c("Russian", "Russian"),
                 lat  = c(55.75, 59.95),
                 long = c(37.616667, 30.3),
# I use here URL shorter by Google
                 urls = c("https://goo.gl/5OUv1E",
                          "https://goo.gl/UWmvDw"))
map.feature(languages = df$lang,
            latitude = df$lat,
            longitude = df$long,
            image.url = df$urls,
            image.width = 150,
            image.height = 150,
            image.X.shift = 10,
            image.Y.shift = 0)
```

Using this argument, users can plot their own markers, any chart connected to a point or even their own legend. It is important to know that by using transparent .png files, the user can plot an additional legend text on the map. 

## 4. Citing `lingtyplogy`
It is important to cite R and R packages when you use them. For this purpose use the `citation` function in R:
```{r}
citation("lingtypology")
```

